---
title: "lungData"
author: "GeZhu"
date: "11/9/2019"
output: html_document
---
## Downloading bioconductor 
Download Bioconductor to local directory
```{r}

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("metagenomeSeq")
```

Download metagenomicSeq locally
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("metagenomeSeq")
```

## Association study pipeline
To load the metagenomeSeq library
```{r}
library(metagenomeSeq)
```

To install biomformat package with BiocManager
```{r}
BiocManager::install(c("biomformat"))
```

Reading in a biom file
```{r}
library(biomformat)
biom_file <- system.file("extdata", "min_sparse_otu_table.biom",
package = "biomformat")
b <- read_biom(biom_file)
biom2MRexperiment(b)
```

## Writing a MRexperiment object out as a BIOM file. 
In this example, we write mouseData MRexperiment object to a BIOM file.You can control the options with normalization.
write_biom is the write function to output object as BIOM file.
```{r}
data(mouseData)
b <- MRexperiment2biom(mouseData)
write_biom(b, biom_file = "˜/Users/gezhu/otu_table.biom")
```

## Loading count data
```{r}
dataDirectory <- system.file("extdata", package = "metagenomeSeq")
lung = loadMeta(file.path(dataDirectory, "CHK_NAME.otus.count.csv"))
dim(lung$counts)
```

## Load annotated taxonomy
Load annotated taxonomy with taxa annotations and OTUs in the same order as matrix rows
```{r}
taxa = read.delim(file.path(dataDirectory, "CHK_otus.taxonomy.csv"),
stringsAsFactors = FALSE)
```

## Loading metadata
load phenotype data into the program with R
```{r}
clin = loadPhenoData(file.path(dataDirectory, "CHK_clinical.csv"),
tran = TRUE)
ord = match(colnames(lung$counts), rownames(clin))
clin = clin[ord, ]
head(clin[1:2, ])
```

## Creating MRexperiment object
We can use function newMRexperiment to takes count matrix, phenoData, and featureData as input. Annotated data frame creation
```{r}
phenotypeData = AnnotatedDataFrame(clin)
phenotypeData
```

annotated taxonomy
```{r}
OTUdata = AnnotatedDataFrame(taxa)
OTUdata
```

```{r}
obj = newMRexperiment(lung$counts,phenoData=phenotypeData,featureData=OTUdata)
obj
```

## Running datasets
In order to run datasets in RStudio with MRexperiment, we need to create MRexperiment object prior to normalization, visualization, and statistical analysis. 
```{r}
data(lungData)
lungData
```

## Access phenotype information
```{r}
phenoData(obj)
head(pData(obj), 3)
#access feature information
featureData(obj)
head(fData(obj)[, -c(2, 10)], 3)
```

Normalize the count matrix
```{r}
head(MRcounts(obj[, 1:2]))
```

Subset MRexperiment-class object
```{r}
featuresToKeep = which(rowSums(obj) >= 100)
samplesToKeep = which(pData(obj)$SmokingStatus == "Smoker")
obj_smokers = obj[featuresToKeep, samplesToKeep]
obj_smokers
head(pData(obj_smokers), 3)
```

NormFactors for alternative normlaization scaling factors
```{r}
head(normFactors(obj))
normFactors(obj) <- rnorm(ncol(obj))
head(normFactors(obj))
```

Library sizes (sequencing depth) can be accessed or replaced with libSize:
```{r}
head(libSize(obj))
libSize(obj) <- rnorm(ncol(obj))
head(libSize(obj))
```

## Filter datasets to a minimum depth/OTU presence
```{r}
data(mouseData)
filterData(mouseData, present = 10, depth = 1000)
```

merge objects with mergeMRexperiment function:
```{r}
data(mouseData)
newobj = mergeMRexperiments(mouseData, mouseData)
newobj
```

## Calculating normalization factors
Normalize data with cumNorm function
```{r}
data(lungData)
p = cumNormStatFast(lungData)
lungData = cumNorm(lungData, p = p)
```

## Calculating normalization factors using Wrench
```{r}
condition = mouseData$diet
mouseData = wrenchNorm(mouseData, condition = condition)
```

## Exporting data
Now we can export the normalized count matrices: 
```{r}
mat = MRcounts(lungData, norm = TRUE, log = TRUE)[1:5, 1:5]
exportMat(mat, file = file.path(dataDirectory, "tmp.tsv"))

```

```{r}
exportStats(lungData[, 1:5], file = file.path(dataDirectory,
"tmp.tsv"))
head(read.csv(file = file.path(dataDirectory, "tmp.tsv"), sep = "\t"))
```

## statistical testing
Detecting differentially abundant features
-Lung microbiome comparison between smoker and non smoker
```{r}
data(lungData)
lungData = lungData[, -which(is.na(pData(lungData)$SmokingStatus))]
lungData = filterData(lungData, present = 30, depth = 1)
lungData <- cumNorm(lungData, p = 0.5)
pd <- pData(lungData)
mod <- model.matrix(~1 + SmokingStatus, data = pd)
lungres1 = fitFeatureModel(lungData, mod)
head(MRcoefs(lungres1))
```

remove unwanted sample
```{r}
data(lungData)
controls = grep("Extraction.Control", pData(lungData)$SampleType)
lungTrim = lungData[, -controls]
rareFeatures = which(rowSums(MRcounts(lungTrim) > 0) < 10)
lungTrim = lungTrim[-rareFeatures, ]
lungp = cumNormStat(lungTrim, pFlag = TRUE, main = "Trimmed lung data")
```

```{r}
lungTrim = cumNorm(lungTrim, p = lungp)
```

Test the bacteria differentially abundant between smokers and non-smokers
```{r}
smokingStatus = pData(lungTrim)$SmokingStatus
bodySite = pData(lungTrim)$SampleType
normFactor = normFactors(lungTrim)
normFactor = log2(normFactor/median(normFactor) + 1)
mod = model.matrix(~smokingStatus + bodySite + normFactor)
settings = zigControl(maxit = 10, verbose = TRUE)
fit = fitZig(obj = lungTrim, mod = mod, useCSSoffset = FALSE,
control = settings)
```

multiple groups (use of limma's toptable function)
```{r}
# maxit=1 is for demonstration purposes
settings = zigControl(maxit = 1, verbose = FALSE)
mod = model.matrix(~bodySite)
colnames(mod) = levels(bodySite)
# fitting the ZIG model
res = fitZig(obj = lungTrim, mod = mod, control = settings)
# The output of fitZig contains a list of various useful
# items. hint: names(res). Probably the most useful is the
# limma 'MLArrayLM' object called fit.
zigFit = slot(res, "fit")
finalMod = slot(res, "fit")$design
contrast.matrix = makeContrasts(BAL.A - BAL.B, OW - PSB, levels = finalMod)
fit2 = contrasts.fit(zigFit, contrast.matrix)
fit2 = eBayes(fit2)
topTable(fit2)
```

exporting fits
```{r}
taxa = sapply(strsplit(as.character(fData(lungTrim)$taxa), split = ";"),
   function(i) {
i[length(i)]
})
head(MRcoefs(fit, taxa = taxa, coef = 2))
```


## Time series analysis
Implement with fitTimeSeries function is the primary method for calculating time intervals of bacteria with different abundance. 
1. log normal permutation test
```{r}
coeffOfInterest = 2
res = fitLogNormal(obj = lungTrim, mod = mod, useCSSoffset = FALSE,
B = 10, coef = coeffOfInterest)
# extract p.values and adjust for multiple testing res$p are
# the p-values calculated through permutation
adjustedPvalues = p.adjust(res$p, method = "fdr")
# extract the absolute fold-change estimates
foldChange = abs(res$fit$coef[, coeffOfInterest])
# determine features still significant and order by the
sigList = which(adjustedPvalues <= 0.05)
sigList = sigList[order(foldChange[sigList])]
# view the top taxa associated with the coefficient of
# interest.
head(taxa[sigList])
## [1] "Veillonella montpellierensis"
## [2] "Veillonella sp. oral clone VeillI7"
## [3] "Listeria grayi"
## [4] "Megasphaera micronuciformis"
## [5] "Prevotella intermedia"
## [6] "Campylobacter curvus"
```

## Presence-absence testing
```{r}
classes = pData(mouseData)$diet
res = fitPA(mouseData[1:5, ], cl = classes)
# Warning - the p-value is calculating 1 despite a high odd's
# ratio.
head(res)
```

## Discovery odds ratio testing
```{r}
classes = pData(mouseData)$diet
res = fitDO(mouseData[1:100, ], cl = classes, norm = FALSE, log = FALSE)
head(res)
```

## Feature correlations
```{r}
cors = correlationTest(mouseData[55:60, ], norm = FALSE, log = FALSE)
head(cors)
```

## Unique OTUs or features
```{r}
cl = pData(mouseData)[["diet"]]
uniqueFeatures(mouseData, cl, nsamples = 10, nreads = 100)
```

## aggregating count 
Normalization at OTU level
```{r}
obj = aggTax(mouseData, lvl = "phylum", out = "matrix")
head(obj[1:5, 1:5])
```

```{r}
obj = aggSamp(mouseData, fct = "mouseID", out = "matrix")
head(obj[1:5, 1:5])
```

## Visualization features
Visualize data sets for metagenomic analysis
```{r}
trials = pData(mouseData)$diet
heatmapColColors = brewer.pal(12, "Set3")[as.integer(factor(trials))]
heatmapCols = colorRampPalette(brewer.pal(9, "RdBu"))(50)

# plotMRheatmap
plotMRheatmap(obj = mouseData, n = 200, cexRow = 0.4, cexCol = 0.4,
trace = "none", col = heatmapCols, ColSideColors = heatmapColColors)
# plotCorr
plotCorr(obj = mouseData, n = 200, cexRow = 0.25, cexCol = 0.25,
trace = "none", dendrogram = "none", col = heatmapCols)
```

## plotting CMDS plots of the data and rarefaction effect at OTU level.
```{r}
cl = factor(pData(mouseData)$diet)
# plotOrd - can load vegan and set distfun = vegdist and use
# dist.method='bray'
plotOrd(mouseData, tran = TRUE, usePCA = FALSE, useDist = TRUE,
bg = cl, pch = 21)
# plotRare
res = plotRare(mouseData, cl = cl, pch = 21, bg = cl)
# Linear fits for plotRare / legend
tmp = lapply(levels(cl), function(lv) lm(res[, "ident"] ~ res[,
"libSize"] - 1, subset = cl == lv))
for (i in 1:length(levels(cl))) {
abline(tmp[[i]], col = i)
}
legend("topleft", c("Diet 1", "Diet 2"), text.col = c(1, 2),
box.col = NA)
27
```

```{r}
head(MRtable(fit, coef = 2, taxa = 1:length(fData(lungTrim)$taxa)))
patients = sapply(strsplit(rownames(pData(lungTrim)), split = "_"),
function(i) {
i[3]
})
pData(lungTrim)$patients = patients
classIndex = list(smoker = which(pData(lungTrim)$SmokingStatus ==
"Smoker"))
classIndex$nonsmoker = which(pData(lungTrim)$SmokingStatus ==
"NonSmoker")
otu = 779
# plotOTU
plotOTU(lungTrim, otu = otu, classIndex, main = "Neisseria meningitidis")
# Now multiple OTUs annotated similarly
x = fData(lungTrim)$taxa[otu]
otulist = grep(x, fData(lungTrim)$taxa)
# plotGenus
plotGenus(lungTrim, otulist, classIndex, labs = FALSE, main = "Neisseria meningitidis")
lablist <- c("S", "NS")
axis(1, at = seq(1, 6, by = 1), labels = rep(lablist, times = 3))

```

```{r}
classIndex = list(Western = which(pData(mouseData)$diet == "Western"))
classIndex$BK = which(pData(mouseData)$diet == "BK")
otuIndex = 8770
# par(mfrow=c(1,2))
dates = pData(mouseData)$date
plotFeature(mouseData, norm = FALSE, log = FALSE, otuIndex, classIndex,
col = dates, sortby = dates, ylab = "Raw reads")
```

Citing metagenomeSeq
```{r}
citation("metagenomeSeq")
```

Session information
```{r}
sessionInfo()
```









